### 2020-02-02 methods.tex
pass over most sections, more to do on sec 5 and 6

\paragraph{Normalization of Empirical Matrices.}  
Normalization is an important, if underappreciated, practical issue.
Importantly, the normalization of weight matrices does \emph{not} affect the Power Law fits because the Heavy Tailed exponent $\alpha$ (as well as other metrics such as the Stable Rank and MP Soft Rank~\cite{MM18_TR,MM19_HTSR_ICML}) is scale-invariant.
Norm-based metrics, however, do depend strongly on the scale of the weight matrix.
Typically, to apply RMT, we would usually define Correlation Matrix with $1/N$ normalization and assume that the variance of $\mathbf{X}$ is either unity or a known constant.% 
\footnote{For formal proofs of Heavy Tailed results, one typically needs a normalization such as $1/N^{1-\alpha}$.}
Pretrained DNNs are typically initialized with random weight matrices $\mathbf{W}_{0}$, with the variance already normalized to $\sqrt{1/N}$, or some variant of this, e.g., the Glorot/Xavier normalization~\cite{GloRot}, or a $\sqrt{2/Nk^2}$ normalization for Convolutional 2D Layers.
We do not have conrol over the final empirical normalization of these models; and we do \emph{not} normalize (or renormalize) the Empirical Correlation Matrices, i.e., we use them as-is.
The only exception to this is that we do rescale the Conv2D pre-Activation Maps $\mathbf{W}_{i,L}$ by $k/\sqrt{2}$ so that they are on the same scale as the Linear layers.

>>>>

\paragraph{Normalization of Empirical Matrices.}  
Normalization is an important, if underappreciated, practical issue.
Importantly, the normalization of weight matrices does \emph{not} affect the Power Law fits because the Heavy Tailed exponent $\alpha$ (as well as other metrics such as the Stable Rank and MP Soft Rank~\cite{MM18_TR,MM19_HTSR_ICML}) is scale-invariant.
Norm-based metrics, however, do depend strongly on the scale of the weight matrix.
Typically, to apply RMT, we would usually define Correlation Matrix with $1/N$ normalization and assume that the variance of $\mathbf{X}$ is either unity or a known constant.% 
\footnote{For Heavy Tailed theorems, one typically needs a normalization such as $1/N^{1-\alpha}$.}
Pretrained DNNs are typically initialized with random weight matrices $\mathbf{W}_{0}$, with the variance already normalized to $\sqrt{1/N}$, or some variant of this, e.g., the Glorot/Xavier normalization~\cite{GloRot}, or a $\sqrt{2/Nk^2}$ normalization for Convolutional 2D Layers.
We do not have conrol over the final empirical normalization of these models; and we do \emph{not} normalize (or renormalize) the Empirical Correlation Matrices, i.e., we use them as-is.
The only exception to this is that we do rescale the Conv2D pre-Activation Maps $\mathbf{W}_{i,L}$ by $k/\sqrt{2}$ so that they are on the same scale as the Linear layers.

### 2020-02-02 nlp_models.tex
pass over most sections, more to do on sec 5 and 6

\charles{More details here.}
The GPT models we analyze are deployed with the popular HuggingFace PyTorch library~\cite{XXX-XXX}.
GPT has 12 layers, with 4 Multi-head Attention Blocks, giving $48$ Layer Weight Matrices $\mathbf{W}$.
Each Block has 2 components, the Self Attention (attn) and the Projection (proj) matrics.  
The self-attention  matrices are larger, of dimension ($2304\times 768$) or ($3072\times 768$).
The projection layer concatenates the self-attention results into a vector (of dimension $768$).
This gives $50$ large matrices.
%
Because GPT and GPT are trained on different data sets, the initial Embedding matrices differ in shape.
GPT  has an initial Token and Positional Embedding layers, of dimension $(40478\times 768)$ and $(512\times 768)$, respectively, whereas GPT2 has input Embeddings of shape $(50257\times 768)$ and $(1024\times 768)$, respectively. 
%
The OpenAI GPT2 (English) models are: \nred{gpt-small, gpt-medium, gpt-large, and gpt-xl}, \
having include $12, 24, 36, \text{and }48$ layers, respectively, with increasingly larger weight matrices.
The model card for GPT2 is published on github.\footnote{\url{https://github.com/openai/gpt-2/blob/master/model_card.md}}.
Table \ref{table:nlp} reports results for the average log norm metrics and the Weighted Alpha metric, 
\michael{Fix that.}

@@

\begin{table}[t]
\small
\begin{center}
%\begin{tabular}{|p{1in}|c|c|c|c|c|}
\begin{tabular}{|p{0.75in}|c|c|c|c|c|}
\hline
%   &    & Frobenius Norm & Spectral Norm & Weighted Alpha & Alpha-Norm \\
% Series & \#Layers   & $\Vert\mathbf{W}\Vert_{F}$ & $\Vert\mathbf{W}\Vert_{\infty}$ & $\hat{\alpha}=\alpha\log\lambda_{max}$ & $\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}$ \\
        &        & Log                   & Log                        & Weighted       & Log                                 \\
        & Num.   & $\Vert\cdot\Vert_{F}$ & $\Vert\cdot\Vert_{\infty}$ & $\hat{\alpha}$ & $\Vert\cdot\Vert^{\alpha}_{\alpha}$ \\
 Series & Layers & Metric                & Metric                     & Metric         & Metric                              \\
\hline
GPT & 49 & 1.64  & 1.72 & 7.01 & 7.28 \\
GPT (small) & 49 & 2.04  & 2.54& 9.62 & 9.87 \\
GPT2 medium & 98 & 2.08 & 2.58& 9.74 & 10.01 \\
GPT2 large & 146 & 1.85 & 1.99& 7.67 & 7.94 \\
GPT2 xl & 194 & 1.86 & 1.92 & 7.17 & 7.51 \\
\hline
\end{tabular}
\end{center}
\caption{Average value for each of the four metrics for pretrainnd OpenAI GPT and GPT2 models.}
\label{table:nlp}
\end{table}

@@

We have analyzed the four quality metrics described in Section~\ref{sxn:methods} for the OpenAI GPT and GPT2 pretrained models.
We start by examining the PL exponents $\alpha$ for GPT and GPT2 (small).
Figure~\ref{fig:GPT-alpha-hist} shows the histogram (empirical density), for all layers, of $\alpha$ for GPT (blue) and this GPT2 (red) model.  
These two histograms are very different, with GPT2 having both a notably smaller mean $\alpha$, and far fewer unusually-large outlying $\alpha$ values.

@@

\begin{figure}
    \centering
    \subfigure[Power Law Exponent $(\alpha)$]{
        %\includegraphics[width=5cm]{img/GPT-alpha-hist.png}
        \includegraphics[width=4.0cm]{img/GPT-alpha-hist.png}
        \label{fig:GPT-alpha-hist}
       }
    %\qquad
    \subfigure[log Spectral Norm $(\log\Vert\mathbf{W}\Vert_{\infty})$]{
        %\includegraphics[width=5cm]{img/GPT-snorm-hist.png}
        \includegraphics[width=4.0cm]{img/GPT-snorm-hist.png}
        \label{fig:GPT-snorm-hist}
       }
   \caption{Comparison of PL exponents, $\alpha$, and log Spectral Norms, $\log\Vert\mathbf{W}\Vert_{\infty}$, for the OpenAI GPT and GPT2 (small) pretrained models.}
\label{fig:GPT-hist}
\end{figure}

@@

    \subfigure[Power Law Exponent $(\alpha)$]{
        %\includegraphics[width=4.5cm]{img/GPT-alpha-depth.png}
        \includegraphics[width=3.7cm]{img/GPT-alpha-depth.png}
        \label{fig:gpt-alpha-layer}
    }
    \qquad
    \subfigure[log Spectral Norm $(\log\Vert\mathbf{W}\Vert_{\infty})$]{
        %\includegraphics[width=4.5cm]{img/GPT-snorm-depth.png}
        \includegraphics[width=3.7cm]{img/GPT-snorm-depth.png}
        \label{fig:gpt-snorm-layer}
    }
    \qquad
    \subfigure[log Alpha-Norm $(\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha})$]{
        %\includegraphics[width=4.5cm]{img/GPT-pnorm-depth.png}
        \includegraphics[width=3.7cm]{img/GPT-pnorm-depth.png}
        \label{fig:gpt-pnorm-layer}
    }
    \caption{Comparison of Correlation Flow and Spectral Norm for OpenAI GPT and GPT2   }
    \label{fig:gpt-alpha-layers}
\end{figure}

@@

We now look across the series of increasingly improving GPT2 models, examining both our
PL exponent $\alpha$, as well as the various Norm metrics.
As we move from small to xl, both the PL exponents $\alpha$, and the peak of distribtion of
log Norm metrics shifts from larger to smaller values. 
Figure \ref{fig:gpt2-histograms} shows the histograms over the layer weight matrics
for fitted $\alpha$, and the log Spectral Norm
 $\log\Vert\mathbf{W}\Vert_{\infty}$  
and log Alpha-Norm
 $\log\Vert\mathbf{W}\Vert_{\alpha}^{\alpha}$ 
metrics.

We see that the average $\alpha$ decreases with increasing model size,
althogh, the differences are less noticible between GTP2 mkodels than between the GPT and GPT2 models.
Unlike GPT, however, the Layer (log) Spectral Norms $\log\vert\mathbf{W}\Vert_{\infty}$ 
and (log) Alpha-Norms $\log\vert\mathbf{W}\Vert_{\alpha}^{\alpha}$
behave more as expected for GPT2 layers, with the larger models  consistently having  smaller norms. 
Likewise, Figure \ref{fig:gpt2-snorm-hist} shows decreasing average log Spectral Norms
with the larger models.   As we have seen in the trends of other well trained models.

We do notice, however, that while the peaks of the $\alpha$ is getting smaller, towards $2.0$,
the tails of the distribution shifts right, with larger GPT2 models having more
usually large $\alpha$.  We suspect this indicates that these larger GPT2 models are 
overparameterized and/or not yet fully optimized and require datasets even larger
than the recent XL $1.5B$ release~\cite{gpt2-xl}.

@@

    \subfigure[Power Law Exponent $(\alpha)$]{
        %\includegraphics[width=4.5cm]{img/GPT2_all_alpha_hist.png}
        \includegraphics[width=3.7cm]{img/GPT2_all_alpha_hist.png}
        \label{fig:gpt2-alpha-hist}
    }
    \qquad
    \subfigure[log Spectral Norm $(\log\Vert\mathbf{W}\Vert_{\infty})$]{
        %\includegraphics[width=4.5cm]{img/GPT2*_all_spectralnorm_hist.png}
        \includegraphics[width=3.7cm]{img/GPT2*_all_spectralnorm_hist.png}
        \label{fig:gpt2-snorm-hist}
    }
    \qquad
    \subfigure[log Alpha-Norm $(\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha})$]{
        %\includegraphics[width=4.5cm]{img/GPT2_all_logpnorm_hist.png}
        \includegraphics[width=3.7cm]{img/GPT2_all_logpnorm_hist.png}
        \label{fig:gpt2-pnorm-hist}
    }
    \caption{Comparison of Power Law Exponnents, Spectral Norm, and Alpha-Norm for different size models in the GPT2 architecture series.  *Note: the log spectral norms (b) histogram omits the first 2 layers, corresponding to the embedding layer, which are normalized differently and have anamolously large values.}
    \label{fig:gpt2-histograms}
\end{figure}

>>>>

\charles{More details here.}
The GPT models we analyze are deployed with the popular HuggingFace PyTorch library~\cite{XXX-XXX}.
GPT has 12 layers, with 4 Multi-head Attention Blocks, giving $48$ Layer Weight Matrices $\mathbf{W}$.
Each Block has 2 components, the Self Attention (attn) and the Projection (proj) matrics.  
The self-attention  matrices are larger, of dimension ($2304\times 768$) or ($3072\times 768$).
The projection layer concatenates the self-attention results into a vector (of dimension $768$).
This gives $50$ large matrices.
%
Because GPT and GPT are trained on different data sets, the initial Embedding matrices differ in shape.
\michael{Clarify that, which is second GPT.}
GPT  has an initial Token and Positional Embedding layers, of dimension $(40478\times 768)$ and $(512\times 768)$, respectively, whereas GPT2 has input Embeddings of shape $(50257\times 768)$ and $(1024\times 768)$, respectively. 
%
The OpenAI GPT2 (English) models are: \nred{gpt-small, gpt-medium, gpt-large, and gpt-xl}, \
having include $12, 24, 36, \text{and }48$ layers, respectively, with increasingly larger weight matrices.
The model card for GPT2 is published on github.\footnote{\url{https://github.com/openai/gpt-2/blob/master/model_card.md}}.

@@

\begin{table}[t]
\small
\begin{center}
%\begin{tabular}{|p{1in}|c|c|c|c|c|}
\begin{tabular}{|p{0.75in}|c|c|c|c|c|}
\hline
%   &    & Frobenius Norm & Spectral Norm & Weighted Alpha & Alpha Norm \\
% Series & \#Layers   & $\Vert\mathbf{W}\Vert_{F}$ & $\Vert\mathbf{W}\Vert_{\infty}$ & $\hat{\alpha}=\alpha\log\lambda_{max}$ & $\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}$ \\
        &        & Log                   & Log                        & Weighted       & Log                                 \\
        & Num.   & $\Vert\cdot\Vert_{F}$ & $\Vert\cdot\Vert_{\infty}$ & $\hat{\alpha}$ & $\Vert\cdot\Vert^{\alpha}_{\alpha}$ \\
 Series & Layers & Metric                & Metric                     & Metric         & Metric                              \\
\hline
GPT & 49 & 1.64  & 1.72 & 7.01 & 7.28 \\
GPT (small) & 49 & 2.04  & 2.54& 9.62 & 9.87 \\
GPT2 medium & 98 & 2.08 & 2.58& 9.74 & 10.01 \\
GPT2 large & 146 & 1.85 & 1.99& 7.67 & 7.94 \\
GPT2 xl & 194 & 1.86 & 1.92 & 7.17 & 7.51 \\
\hline
\end{tabular}
\end{center}
\caption{Average value for the Average Log Norm metrics and the Weighted Alpha metric for pretrainnd OpenAI GPT and GPT2 models.  }
\label{table:nlp}
\end{table}

@@

We have analyzed the four quality metrics described in Section~\ref{sxn:methods} for the OpenAI GPT and GPT2 pretrained models.
See Table \ref{table:nlp} for a summary of results.
We start by examining the PL exponents $\alpha$ for GPT and GPT2 (small).
Observe that all four metrics increase when going from GPT to GPT (small), i.e., they are smaller for the higher-quality model (higher quality since GPT was trained to better data), when the number of layers is held fixed.
Observe also that (with one minor exception involving the Log Frobenius norm metric) all four metrics decrease as one goes from GPT2 medium to GPT2 large to GPT2 xl, indicating that the larger models are indeed better than the smaller models.
%(The one minor exception is the Log Frobenius norm increasing from $1.85$ to $1.86$, for GPT2 large to GPT2 xl.)
\michael{We seem to have typos(s), so I'm not going to flesh out all conclusions I see yet.  The comments I just mad may be wrong, if GPT (small) is really GPT2 (small).  FIX TYPOS AND CLARIFY.}
These results are broadly consistent with the results in Section~\ref{sxn:cv} on series of CV models.

Figure~\ref{fig:GPT-alpha-hist} shows the histogram (empirical density), for all layers, of $\alpha$ for GPT (blue) and this GPT2 (red) model.  
These two histograms are very different, with GPT2 having both a notably smaller mean $\alpha$, and far fewer unusually-large outlying $\alpha$ values.

@@

\begin{figure}
    \centering
    \subfigure[PL exponent ($\alpha$)]{
        %\includegraphics[width=5cm]{img/GPT-alpha-hist.png}
        \includegraphics[width=4.0cm]{img/GPT-alpha-hist.png}
        \label{fig:GPT-alpha-hist}
       }
    %\qquad
    \subfigure[Log Spectral Norm]{
        %\includegraphics[width=5cm]{img/GPT-snorm-hist.png}
        \includegraphics[width=4.0cm]{img/GPT-snorm-hist.png}
        \label{fig:GPT-snorm-hist}
       }
   \caption{Comparison of PL exponents, $\alpha$, and log Spectral Norms, $\log\Vert\mathbf{W}\Vert_{\infty}$, for the OpenAI GPT and GPT2 (small) pretrained models.}
\label{fig:GPT-hist}
\end{figure}

@@

    \subfigure[PL exponent ($\alpha$)]{
        %\includegraphics[width=4.5cm]{img/GPT-alpha-depth.png}
        \includegraphics[width=4.0cm]{img/GPT-alpha-depth.png}
        \label{fig:gpt-alpha-layer}
    }
    %\qquad
    \subfigure[Log Spectral Norm]{
        %\includegraphics[width=4.5cm]{img/GPT-snorm-depth.png}
        \includegraphics[width=4.0cm]{img/GPT-snorm-depth.png}
        \label{fig:gpt-snorm-layer}
    }
    %\qquad
    \subfigure[Log Alpha Norm]{
        %\includegraphics[width=4.5cm]{img/GPT-pnorm-depth.png}
        \includegraphics[width=4.0cm]{img/GPT-pnorm-depth.png}
        \label{fig:gpt-pnorm-layer}
    }
    \caption{Comparison of Correlation Flow and Spectral Norm for OpenAI GPT and GPT2   }
    \label{fig:gpt-alpha-layers}
\end{figure}

@@

We now look across the series of increasingly improving GPT2 models, examining both the PL exponent $\alpha$ as well as the various Log Norm metrics.
In general, as we move from small to xl, the histograms for both $\alpha$ exponents and the log Norm metrics shift from larger to smaller values. 
See Figure \ref{fig:gpt2-histograms}, which shows the histograms over the layer weight matrics
for fitted $\alpha$, and the log Spectral Norm
 $\log\Vert\mathbf{W}\Vert_{\infty}$  
and log Alpha Norm
 $\log\Vert\mathbf{W}\Vert_{\alpha}^{\alpha}$ 
metrics.

We see that the average $\alpha$ decreases with increasing model size, although the differences are less noticible between different GTP2 models than between the GPT and GPT2 models.
Unlike GPT, however, the Layer (log) Spectral Norms $\log\vert\mathbf{W}\Vert_{\infty}$ 
and (log) Alpha Norms $\log\vert\mathbf{W}\Vert_{\alpha}^{\alpha}$
behave more as expected for GPT2 layers, with the larger models  consistently having  smaller norms. 
\michael{Clarify.}
Likewise, Figure \ref{fig:gpt2-snorm-hist} shows decreasing average log Spectral Norms with the larger models.  
As we have seen in the trends of other well trained models.

We do notice, however, that while the peaks of the $\alpha$ is getting smaller, 
towards $2.0$,
the tails of the distribution shifts right, with larger GPT2 models having more
usually large $\alpha$.  
We suspect this indicates that these larger GPT2 models are over-parameterized/under-optimized and 
that they have capacity to support datasets even larger than the recent XL $1.5B$ release~\cite{gpt2-xl}.

@@

    \subfigure[PL exponent ($\alpha$)]{
        %\includegraphics[width=4.5cm]{img/GPT2_all_alpha_hist.png}
        \includegraphics[width=4.0cm]{img/GPT2_all_alpha_hist.png}
        \label{fig:gpt2-alpha-hist}
    }
    %\qquad
    \subfigure[Log Spectral Norm]{
        %\includegraphics[width=4.5cm]{img/GPT2*_all_spectralnorm_hist.png}
        \includegraphics[width=4.0cm]{img/GPT2*_all_spectralnorm_hist.png}
        \label{fig:gpt2-snorm-hist}
    }
    %\qquad
    \subfigure[Log Alpha Norm]{
        %\includegraphics[width=4.5cm]{img/GPT2_all_logpnorm_hist.png}
        \includegraphics[width=4.0cm]{img/GPT2_all_logpnorm_hist.png}
        \label{fig:gpt2-pnorm-hist}
    }
    \caption{PL exponents ($\alpha$), Log Spectral norm, and Log Alpha norm for different size models in the GPT2 architecture series.  (The Log Spectral norm histogram omits the first 2 layers, corresponding to the embedding layer, as these are normalized differently and have anamolously large values.)}
    \label{fig:gpt2-histograms}
\end{figure}
